---
title: "Introduction to programming for data analysis"
subtitle: "Session 4 — Statistical Tests and Regressions"
author:
  - name: Ulrich Aiounou
    email: dimitri.aiounou@univ-amu.fr
    url: https://research.ulrichaiounou.com/fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Ewen Gallic
    email: ewen.gallic@univ-amu.fr
    url: https://egallic.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Pierre Michel
    email: pierre.michel@univ-amu.fr
    url: https://www.amse-aixmarseille.fr/fr/membres/michel
    affiliations:
      - name: AMSE, Aix Marseille University
format:
  revealjs:
    theme: [simple, ../assets/custom.scss]
    logo: "figs/logo_AMSE_noir_RVB.png"
    slide-number: true
    transition: fade
    preview-links: true
    footer: "Introduction to programming for data analysis — Session 2"
    code-copy: true
    code-line-numbers: true
    highlight-style: github
    smaller: false
    html-math-method: mathjax
    include-in-header:
      - file: ../assets/mathjax-color.html
      - file: ../assets/_homepage-link.html
      - text: |
          <link rel="stylesheet"
                href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/6.5.2/css/all.min.css">
execute:
  echo: true
  warning: false
  message: false
  fig-width: 9
  fig-height: 5.5
  fig-align: "center"
  dev: "svg" # Clean vector output for ggplot2 plots
---



## Disclaimer {background-color="#F5F6F0"}

$$
\definecolor{wongBlack}{RGB}{0,0,0}
\definecolor{wongGold}{RGB}{230, 159, 0}
\definecolor{wongLightBlue}{RGB}{86, 180, 233}
\definecolor{wongGreen}{RGB}{0, 158, 115}
\definecolor{wongYellow}{RGB}{240, 228, 66}
\definecolor{wongBlue}{RGB}{0, 114, 178}
\definecolor{wongOrange}{RGB}{213, 94, 0}
\definecolor{wongPurple}{RGB}{204, 121, 167}
\definecolor{colA}{RGB}{255, 221, 85}
\definecolor{colB}{RGB}{148, 78, 223}
\definecolor{colC}{RGB}{63, 179, 178}
\definecolor{colGpeZero}{RGB}{127, 23, 14}
\definecolor{colGpeUn}{RGB}{27, 149, 224}
$$

* The slides were adapted from those of Pierre Michel and Morgan Raux, both researchers at AMSE, who kindly shared their work.

* This slide deck was made with Quarto and reveal.js, it is translated by Claude Sonnet 5 from a LaTex presentation previously made with beamer.



## The 4 steps in any data analysis

1. Finding data
2. Cleaning data to combine several data sources
3. **Producing knowledge out from data (summary statistics, regression analysis summarised via tables and graphs)**
4. Interpreting the results (writing)

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(modelr)
library(forcats)
library(tidyverse)
```

# Statistical Hypothesis Tests {.part}

## What is a statistical hypothesis test?

- A <span class="hl-blue">statistical hypothesis test</span> is a statistical method for deciding between two hypotheses.
- It consists in rejecting (or not) a statistical hypothesis, called the <span class="hl-purple">null hypothesis</span>.
- The statistic of the test is assumed to follow a specific distribution under the null hypothesis.
- Using <span class="hl-blue">historical data</span>, an <span class="hl-purple">observed value</span> for that statistic is computed.
- Then, a comparison with what we would expect if the null were true is made, based on the p-value:
  - this comparison allows us to decide whether the difference is likely due to chance (in this case, the null is not rejected) or not.

## Example: Student's t-Test

A common hypothesis test is the <span class="hl-blue">Student's t-Test</span>, used to compare the means of two independent samples.

- Consider two random samples of observations from two distributions $P_1$ and $P_2$, with population means $\mu_1$ and $\mu_2$.
- Let $\overline{X}_1$ and $\overline{X}_2$ be the sample means (with observed values $\overline{x}_1$ and $\overline{x}_2$).
- We test whether the observed difference $\overline{x}_1 - \overline{x}_2$ is statistically distinguishable from 0 to infer whether $\mu_1 = \mu_2$.
- In other words, the t-test is about testing the null hypothesis:
$$\text{H}_0: \mu_1 = \mu_2$$
using the observed difference $\overline{x}_1 - \overline{x}_2$, while accounting for sample variability.

## Example: Student's t-Test

- The T-statistic is
$$
T = \frac{\overline{X}_1 - \overline{X}_2}{\sqrt{\tfrac{s_1^2}{n_1} + \tfrac{s_2^2}{n_2}}},
$$
where $s_1^2, s_2^2$ are the sample variances and $n_1, n_2$ the sample sizes.
- Under $\text{H}_0$, $T$ approximately follows a Student's $t$-distribution with $n-1$ degrees of freedom.

## Example: Student's t-Test

- The <span class="hl-blue">decision rule</span> for the t-Test (two-sided test) is as follows:
  - We reject $\text{H}_0$ if $|T| > t_{1-\alpha/2, \, \nu}$,
  - where $t_{1-\alpha/2, \, \nu}$ is the critical value of the $t$-distribution with $\nu$ degrees of freedom,
  - and where $\alpha$ is the significance level (usually $5\%$).
- Equivalently, we reject $\text{H}_0$ if the p-value $< \alpha$
  - The p-value is the probability, under $H_0$, of observing a test statistic at least as extreme as the one computed from the data, under the null:
$$
p = \mathbb{P}\big[ \mid T_\nu \mid \geq \mid t_{\text{obs}}\mid \big] = 2 \times \mathbb{P}\big[ T_\nu \geq \mid t_{\text{obs}}\mid\big],
$$

## Application to GDP and life satisfaction

- [We want to test whether the average GDP of 'happy' countries differs significantly from the average GDP of 'unhappy' countries (in 2018).]{.gras-secondCol}
- We assume that:
  - we have independent observations across countries,
  - each group's sampling distribution of the mean is approximately normal (if $n>30$ for example, which is not the case here...).

## Application to GDP and life satisfaction

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
tb <- read_csv("data/out/gdp_lifesatisf.csv")
happy <- filter(tb, 
  life_satisf >= 7, year == 2018
  )
unhappy <- filter(
  tb, life_satisf < 7, year == 2018
  )
t.test(happy$gdp, unhappy$gdp)
```
:::
::: {.column width="50%"}
```{r}
#| echo: false
tb <- read_csv("../tutorials/data/out/gdp-lifesat.csv")
happy <- filter(tb, 
  life_satisf >= 7, year == 2018
  )
unhappy <- filter(
  tb, life_satisf < 7, year == 2018
  )
t.test(happy$gdp, unhappy$gdp)
```


- What do you conclude?
:::
::::

# Linear Regression {.part}

## Linear Regression

- Recall that a linear regression is used to model the relationship between a <span class="hl-blue">dependent variable</span> (usually denoted <span class="hl-blue">$Y$</span>) and one or more <span class="hl-purple">explanatory variables</span> (usually denoted by <span class="hl-purple">$X_1, X_2, \ldots, X_p$</span>).
- Here, let us focus on the simple linear regression, with only one variable $X\in \mathcal{X}$.
- We assume that the relationship between the <span class="hl-blue">dependent variable $Y$</span> and the <span class="hl-purple">independent variable $X$</span> is linear:
$$
{\color{blue}Y} = a + b{\color{purple}X}
$$

## Example

- Going back to our GDP/Life satisfaction example, assume that the <span class="hl-blue">response variable $Y$</span> is life satisfaction and the <span class="hl-purple">explanatory variable $X$</span> is GDP.
- $Y = a + bX$, where:
  - $a$ is the <span class="hl-green">intercept (constant)</span>,
  - $b$ is the <span class="hl-purple">slope</span> of the line.
- The values $a$ and $b$ are not known beforehand.

## Example

:::: {.columns}
::: {.column width="60%"}
```{r}
#| eval: false
ggplot(
  data = tb |> filter(year == 2018),
  mapping = aes(x = gdp, y = life_satisf)
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Real per capita GDP in 2013 (2020 Euro)",
    y = "Life satisfaction"
  ) +
  scale_x_continuous(
    labels = scales::label_number(
      suffix = "", scale = 1, big.mark = ","
    )
  )
```
:::
::: {.column width="40%"}
```{r}
#| echo: false
#| fig-cap: "Relationship between life satisfaction and per capita GDP, with OLS regression line (in blue)."
ggplot(
  data = tb |> filter(year == 2018),
  mapping = aes(x = gdp, y = life_satisf)
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Real per capita GDP in 2013 (2020 Euro)",
    y = "Life satisfaction"
  ) +
  scale_x_continuous(
    labels = scales::label_number(
      suffix = "", scale = 1, big.mark = ","
    )
  )
```

:::
::::

## The Regression Model

- In the previous slide, we assumed the following model:
$$
{\color{blue}Y_i} = {\color{green}\beta_0} + {\color{orange}\beta_1} {\color{purple}X_i} +\varepsilon_i,
$$
where

  - $i=1,\ldots,n$ represent the individual, where $n$ is the total number of individuals,
  - <span class="hl-blue">$Y_i$</span> is the <span class="hl-blue">dependent variable</span>,
  - <span class="hl-purple">$X_i$</span> is the <span class="hl-purple">independent variable</span>,
  - <span class="hl-green">$\beta_0$</span> is the <span class="hl-green">intercept</span> of the regression line,
  - <span class="hl-orange">$\beta_1$</span> is the <span class="hl-orange">slope</span> of the regression line,
  - $\varepsilon_i$ is the error, where we assume $\varepsilon \sim \mathcal{N}\big(0, \sigma^2\big)$.

## The OLS Estimator

- The coefficients <span class="hl-green">$\beta_0$</span> and <span class="hl-orange">$\beta_1$</span> can be estimated so as to minimize the sum of squared residuals (ordinary least squares (OLS) estimator):
$$
(\hat{\beta}_0, \hat{\beta}_1)= \text{arg}\min_{\beta_0, \beta_1} \sum_{i=1}^{n}(Y_i - \beta_0 - \beta_1 X_i)^2
$$
- Graphically speaking, this corresponds to finding the values of $\beta_0$ and $\beta_1$ that will produce a line that is "as close as possible to the observations."
- You saw in your regression class that in the case of simple linear regression, we have:
  - $\hat{\beta}_0 = \overline{Y}-\hat{\beta}_1 \overline{X}$
  - $\hat{\beta}_1 = \frac{\sum_{i=1}^{n} (X_i - \overline{X})(Y_i - \overline{Y})}{\sum_{i=1}^{n}(X_i-\overline{X})^2}$

## Predicted Values and Residuals

- The predicted values using the OLS estimator, denoted $\hat{y}_i$, are computed as:
$$
\hat{y}_i =\hat{\beta}_0 + \hat{\beta}_1 X_i
$$
- The distance between the observed value $Y_i$ and the predicted value $\hat{y}_i$ is named the residual, denoted $\hat{\varepsilon}$:
$$
\hat{\varepsilon}_i = Y_i - \hat{y}_i
$$

::: {.tip}
<span class="box-title">Interactive Application</span>
A cool interactive online graphical application to understand the mechanisms of OLS can be accessed at the following URL: [https://www.econometrics-with-r.org/SimpleRegression.html](https://www.econometrics-with-r.org/SimpleRegression.html)
:::

## A Linear Regression with R

- In `R`, the OLS estimators of a linear model can be easily obtained, using the `lm()`{.R} function:
  - we provide the `lm()`{.R} function with a formula of the relationship we want to model,
  - and a dataset.
- For example, in the dataset `sim1` included in the `modelr` package, there are two variables, `y` and `x`. If we want to explain the variations of `y` by the variations of `x`, we write:

```{r}
mod_sim1 <- lm(formula = y ~ x, data = modelr::sim1)
```

## The Estimated Model

- When we print the object returned by `lm()`{.R} in the console, we have:

```{r}
mod_sim1
```

Hence, the estimated model writes:
$$
\hat{y}_i = 4.221 + 2.05 \times x_i
$$

## A Linear Regression with R: Estimates

- The coefficients of the model, estimated with OLS, can be accessed using the `coef()`{.R} function.

```{r}
coef(mod_sim1)
coef(mod_sim1)["(Intercept)"]
coef(mod_sim1)["x"]
```

## A Linear Regression with R: Predictions

Instead of computing the predicted values using the extracted values of the coefficients and the matrix of explanatory variables, `R` offers the function `predict()`{.R}:

```{r}
predict(mod_sim1)
```

## A Linear Regression with R: Residuals

Similarly, for the residuals, `R` provides the `residuals()`{.R} function:

```{r}
residuals(mod_sim1)
```

## A Linear Regression with R: Summary

The `summary()`{.R} function applied to a linear regression model provides a lot of useful information on model fit:

```{r}
summary(mod_sim1)
```

## Export Your Regression Table {.smaller}

- Please do not make a screenshot of a regression table from your statistical software.
- In `R`, there is a convenient package called `{stargazer}` which provides a function with the same name that allows you to export your summary table as a nice table in HTML or LaTeX format.
  - If you write your report using Google Docs, MS Word, OpenOffice, or Pages, export your table in HTML. You will then be able to copy/paste the table from your browser into your document (change the `type` argument).

:::{.columns}

:::{.column width="50%"}
```{r}
#| eval: false
library(stargazer)
stargazer(mod_sim1, type = "text")
stargazer(mod_sim1, type = "html")
stargazer(mod_sim1, type = "latex")
```

Using `type = "text"`{.R} (as shown on the right-hand side) is convenient for looking at your results in the console. However, in a report, prefer another format.

:::

:::{.column width="50%"}

```{r}
#| echo: false
#| message: false
#| warning: false
library(stargazer)
stargazer(mod_sim1, type = "text")
```
:::

:::

# Visualize Predicted Values {.part}

## Predictions

:::: {.columns}
::: {.column width="60%"}
```{r}
#| eval: false
ggplot(
  data = sim1 |> mutate(
    pred = predict(
      mod_sim1, newdata = sim1
    )
  )
) +
  geom_point(mapping = aes(x = x, y = y)) +
  geom_line(
    mapping = aes(x = x, y = pred),
    colour= "dodgerblue", 
    linewidth = 1
  )
```

The blue line traces the predicted values from `mod_sim1` across the range of `x`.
:::
::: {.column width="40%"}
```{r}
#| echo: false
ggplot(
  data = sim1 |> mutate(
    pred = predict(
      mod_sim1, newdata = sim1
    )
  )
) +
  geom_point(mapping = aes(x = x, y = y)) +
  geom_line(
    mapping = aes(x = x, y = pred),
    colour= "dodgerblue", 
    linewidth = 1
  )
```
:::
::::

## Residuals

:::: {.columns}
::: {.column width="50%"}
```{r}
ggplot(
  data = sim1 |> mutate(
    resid = residuals(mod_sim1)
  )
) +
  geom_histogram(
    mapping = aes(x = resid),
    binwidth = 0.5, colour = "white"
  ) +
  labs(x = "Residuals", y = "Count")
```
:::
::: {.column width="50%"}
The distribution of residuals should look roughly symmetric and centered around 0 if the model fits well.
:::
::::

## Quantile-Quantile Plot

- A <span class="hl-purple">QQ plot</span> (<span class="hl-purple">quantile–quantile plot</span>) in linear regression compares the sorted residuals from the model to the <span class="hl-blue">theoretical quantiles of a normal distribution</span>.
- This provides a visualization to <span class="hl-blue">check the normality assumption of residuals</span>.
  - If the residuals are normally distributed, they will roughly lie on the bisector.
  - If we observe an S-shape pattern, this points to a skewness of the distribution of the residuals.
  - If we observe curved tails, it implies heavy or light tails for the distribution of the residuals.
  - If we observe isolated endpoints, this is a hint that there may be outliers.

## Residuals: Quantile-Quantile Plot

:::: {.columns}
::: {.column width="50%"}
```{r}
ggplot(
  data = sim1 |> mutate(
    resid = residuals(mod_sim1)
  ),
  mapping = aes(sample = y)
) + stat_qq() + stat_qq_line() +
  labs(title = "Normal QQ plot")
```
:::
::: {.column width="50%"}
Points close to the diagonal line indicate residuals consistent with a normal distribution.
:::
::::

## Residuals: Relationship with $X$?

:::: {.columns}
::: {.column width="50%"}
```{r}
ggplot(
  data = sim1 |> mutate(
    resid = residuals(mod_sim1)
  )
) +
  geom_ref_line(h = 0) +
  geom_point(
    mapping = aes(x = x, y = resid),
  ) +
  labs(y = "Residuals")
```
:::
::: {.column width="50%"}
No clear pattern between `x` and the residuals is a good sign. A visible trend would suggest a missing non-linear term.
:::
::::

# Regression with categorical features {.part}

## Reference Class

- A character variable included in a regression will be transformed into a `factor` in `R`.
- By default, the reference class is the first level in alphanumerical order.
- To choose a different reference category:
  - specify the order manually with the `levels` argument of the `factor()`{.R} function,
  - or use `fct_relevel()`{.R} from the `{forcats}` package to directly set the reference level.
- The reference level determines the baseline against which other categories are compared in the regression.

## Reference Class: Benchmark

- Benchmark situation: the variable `x` is a character variable which contains the following unique values: `"a"`{.R}, `"b"`{.R}, `"c"`{.R}, `"d"`{.R}.
- The reference category will thus be `"a"`{.R}.

```{r}
lm(y ~ x, data = sim2)
```

## Changing the Reference Class

Using the `factor()`{.R} function or the `fct_relevel()`{.R} function from `{forcats}`:

:::{.panel-tabset}

### `factor()`{.R}

```{r}
lm(
  y ~ x,
  data = sim2 |> 
    mutate(x = factor(x, levels = c("c", "a", "b", "d"))),
)
```

### `fct_relevel()`{.R}

```{r}
lm(
  y ~ x,
  data = sim2 |> 
    mutate(x = fct_relevel(x, "c"))
)
```
:::



## Bivariate Graph for Residuals

:::: {.columns}
::: {.column width="50%"}
```{r}
mod_sim2 <- lm(y ~ x, data = sim2)
ggplot(
  data = sim2 |> 
    mutate(pred = predict(mod_sim2)),
  mapping = aes(x = x)
)+
  geom_point(mapping = aes(y = y)) +
  geom_point(
    mapping = aes(y = pred),
    colour = "red"
  )
```
:::
::: {.column width="50%"}
`R` automatically converts categorical variables included in a model.
:::
::::

# Interaction Terms {.part}

## Interaction in a Model

We may believe the effect of a variable ($x_1$) depends on the value of another ($x_2$).

:::: {.columns}
::: {.column width="50%"}
```{r}
# the sim3 dataset is from {modelr}
ggplot(
  data = sim3, 
  mapping = aes(
    x = x1, y = y, colour = x2)
  ) +
  geom_point() +
  geom_smooth()
```
:::
::: {.column width="50%"}
The slope of `y` with respect to `x1` appears to differ across levels of `x2`.
:::
::::

## Interaction in a Model

- Benchmark model without interactions:
$$
y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + \varepsilon_i, 
\quad \varepsilon_i \sim \mathcal{N}(0, \sigma^2).
$$
- To test whether the effect of $x_1$ depends on the value of $x_2$, we add an interaction term:
$$
y_i = \gamma_0 + \gamma_1 x_{1,i} + \gamma_2 x_{2,i} 
      + \gamma_3 (x_{1,i} \times x_{2,i}) + \varepsilon_i, 
\quad \varepsilon_i \sim \mathcal{N}(0, \sigma^2).
$$
- With this specification, the marginal effect of $x_1$ is
$$
\frac{\partial y_i}{\partial x_{1,i}} = \gamma_1 + \gamma_3 x_{2,i},
$$
Hence, the slope of $x_1$ changes with $x_2$.

## Interaction in a Model

We will consider two cases:

1. a continuous variable interacted with a <span class="hl-blue">categorical variable</span>,
2. a continuous variable interacted with a <span class="hl-blue">continuous variable</span>.

## Interaction Between Continuous and Categorical Variables

```{r}
(mod_bench <- lm(y ~ x1 + x2, data = sim3))

(mod_interact <- lm(y ~ x1 * x2, data = sim3))
```

## Visualize the Slopes in a Graph

```{r}
#| output-location: slide
#| code-line-numbers: "3,13,14,15,16,17,18"
ggplot(
  data = sim3 |> 
    modelr::gather_predictions(mod_bench, mod_interact) |> 
    mutate(
      model = factor(
        model, 
        levels = c("mod_bench", "mod_interact"), 
        labels = c("Benchmark", "Interaction~between~x[1]~and~x[2]"))
    ),
  mapping = aes(x = x1, y = y, colour = x2)
) +
  geom_point() +
  geom_line(aes(y = pred)) +
  facet_wrap(
    ~ model, 
    labeller = labeller(
      model = label_parsed
    )
)
```

## A Quick Glance at the Residuals

```{r}
#| output-location: slide
#| code-line-numbers: "10"
ggplot(
  data = sim3 |> 
    modelr::gather_residuals(mod_bench, mod_interact) |> 
    mutate(
      model = factor(
        model, 
        levels = c("mod_bench", "mod_interact"), 
        labels = c("Benchmark", "Interaction~between~x[1]~and~x[2]"))
    ),
  mapping = aes(x = x1, y = resid, colour = x2)
) +
  geom_point() +
  facet_grid(
    model ~ x2,
    labeller = labeller(
      model = label_parsed
    )
  ) +
  labs(y = "residuals")
```

## Interaction Between Two Continuous Variables

```{r}
(mod1 <- lm(y ~ x1 + x2, data = sim4))

(mod2 <- lm(y ~ x1 * x2, data = sim4))
```

## Interaction Between Two Continuous Variables

```{r}
modelr::gather_predictions(data = sim4, mod1, mod2)
```

## Interaction Between Two Continuous Variables

```{r}
#| output-location: slide
ggplot(
  data = sim4 |> 
    modelr::gather_predictions(mod1, mod2) |> 
    mutate(
      model = factor(
        model, 
        levels = c("mod1", "mod2"), 
        labels = c("Benchmark", "Interaction~between~x[1]~and~x[2]"))
    ),
  mapping = aes(x = x1, y = pred, color = x2, group = x2)
) +
  geom_line() + 
  facet_wrap( ~ model)
```

# Non-Linear Terms in a Linear Regression Model {.part}

## Non-Linear Terms

- The relationship with an explanatory variable and the response variable may be non-linear.
- Introducing transformed variables into the model allows us to take these non-linearities into account.
  - log, squared value, power
- In an `R` formula, if the transformation involves one of the following signs, the transformation must appear inside the ``I()``{.R} function when writing the formula:
  - `+`{.R}, `*`{.R}, `^`{.R}, `-`{.R}.

## Example

- For example, assume we want to fit the following model:
$$
y_i = \beta_0 + \beta_1 x_i + \beta_2x_i^2 + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0,1).
$$
- The following <span class="hl-blue">formula</span> will be used:

```{r}
y ~ x + I(x^2)
```

- Alternatively, one can use the `poly()`{.R} function (polynomials):

```{r}
y ~ poly(x, 2, raw = TRUE)
```

# Regression with Panel Data {.part}

## Panel Data

- In session 1, we mentioned panel data.
- Recall that <span class="hl-purple">cross-sectional data</span> contain observations that are collected at a single point in time across multiple units.
- In contrast, <span class="hl-purple">panel data</span> contain observations for multiple units followed over two or more time periods.
- A panel dataset will be denoted as follows:
$$\{ (\boldsymbol{X}_{it}, y_{it}) \}_{i=1,\dots,n;\; t=1,\dots,T},$$
where $i=1,\ldots n$ refers to **individuals** and $t=1,\ldots$ refers to the **time period**.

## Example

The dataset containing real per capita GDP merged with the dataset giving life satisfaction, both downloaded from Eurostat (see [tutorial 1](tutorial-1.html)) contains annual values (from 2010 to 2023) for multiple countries (6).
```{r}
tb
```

## Panel Data Setup

- We have a panel dataset: $N=6$ countries observed over $T=14$ years (even if it is not perfectly balanced).
- The variables:
  - $gdp_{it}$: per capita GDP,
  - $life_{it}$: life satisfaction (response variable).
- Our regression model of interest:
$$
\text{life\_satisf}_{it} = \beta_0 + {\color{wongGreen}\beta_1} \text{gdp}_{it} + \varepsilon_{it}.
$$
- How can we account for <span class="hl-orange">shocks common to all countries in the same year</span> (e.g., financial crisis, pandemic)?

## Why Time Fixed Effects?

- The <span class="hl-orange">events</span> mentioned previously (global recessions, pandemics, or geopolitical crises) affect all countries simultaneously.
- Ignoring these can bias $\color{wongGreen}\beta_1$ if GDP and <span class="hl-orange">these shocks</span> are correlated.
- To tackle this issue, we add **time dummies** ${\color{wongOrange}\delta_t}$ to capture common <span class="hl-orange">year-specific effects</span>.

## Panel Regression with Time Fixed Effects

- Extended specification:
$$
\text{life\_satisf}_{it} = \beta_0 + {\color{wongGreen}\beta_1} \text{gdp}_{it} + \sum_{t\in\mathcal{T}} {\color{wongOrange}\delta_t} D_t + u_{it},
$$
where $D_t$ is a dummy variable equal to 1 if year = $t$, and $\mathcal{T} = \{2013, 2018, 2021, 2022, 2023, 2024\}$.
- We interpret the coefficients as follows:
  - ${\color{wongGreen}\beta_1}$: marginal effect of per capita GDP on life satisfaction, <span class="hl-blue">net of year-specific shocks</span>.
  - $\color{wongOrange}\delta_t$: average shift in life satisfaction in year $t$ relative to the baseline year.

## Implementation in R: Pooled OLS

The FE model estimated with the `lm()`{.R} function:

```{r}
#| eval: false
mod_fe <- lm(
  life ~ gdp + factor(year),
  data = panel_data
)
```

Using the `{plm}` package:

```{r}
#| eval: false
library(plm)
tb_pdata <- pdata.frame(tb, index = c("country", "year"))
mod_fe_plm <- plm(
  life_satisf ~ gdp + factor(year),
  data = tb_pdata,
  model = "pooling"
)
```


## With Country Fixed Effects

- So far, we controlled for <span class="hl-orange">time fixed effects</span> $\delta_t$ (global shocks common to all countries).
- But countries also differ in <span class="hl-gold">time-invariant characteristics</span>, for example:
  - culture, geography, history,
  - institutions, long-run policies, etc.
- These may be correlated with per capita GDP and life satisfaction.
- Hence, to account for this, we can also control for <span class="hl-gold">country fixed effects</span> $\alpha_i$.

## Two-Way Fixed Effects Model (1/2)

- The regression equation becomes:
$$
\text{life\_satisf}_{it} = \beta_0 + {\color{wongGreen}\beta_1} \text{gdp}_{it} + {\color{wongGold}\alpha_i} + {\color{wongOrange}\delta_t} + \varepsilon_{it},
$$
  - $\color{wongGold}\alpha_i$: country-specific effects (to absorb unobserved, time-invariant differences across countries).
  - $\color{wongOrange}\delta_t$: time effects (to capture common shocks in a given year).

## Two-Way Fixed Effects Model (2/2)

- $\color{wongGreen}\beta_1$ measures the effect of GDP on life satisfaction <span class="hl-blue">within countries over time</span>, net of:
  - <span class="hl-gold">persistent cross-country differences</span> ($\color{wongGold}\alpha_i$),
  - <span class="hl-orange">global shocks or year-specific changes</span> ($\color{wongOrange}\delta_t$).
- In other words:
  - we compare each country to *itself* over time,
  - while adjusting for events that affect all countries in the same year.

## Implementation in R

With the `lm()`{.R} function:

```{r}
#| eval: false
mod_twfe_lm <- lm(
  life_satisf ~ gdp + factor(country) + factor(year),
  data = tb
)
```

With the `plm()`{.R} function:

```{r}
#| eval: false
mod_twfe <- plm(
  life_satisf ~ gdp,
  data = tb_pdata,
  model = "within",
  effect = "twoways"
)
```


## Package for Estimations with Multiple Fixed-Effects

The package `{fixest}` seems nice:

- The homepage of the package: [🌐 fixest: Fast and user-friendly fixed-effects estimation](https://lrberge.github.io/fixest/)
- A cool vignette for the package: [🌐 Fast Fixed-Effects Estimation: Short Introduction](https://cran.r-project.org/web/packages/fixest/vignettes/fixest_walkthrough.html)

```{r}
#| eval: false
library(fixest)
mod_fe_fixest <- feols(life_satisf ~ gdp | year, data = tb)
etable(mod_fe_fixest)

mod_twfe_fixest <- feols(life_satisf ~ gdp | year + country, data = tb)
etable(mod_twfe_fixest)
```

## Exercises {.exercise}

Practice with [the third tutorial](tutorial-3.html)!
