---
title: "Quick Recap: Creating an RStudio Project"
subtitle: "Introduction to programming for data analysis"
author:
  - name: Ewen Gallic
    email: ewen.gallic@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
format:
  html:
    css: styles.css
    include-in-header: _fonts.html
    include-before-body: _topbar.html
    include-after-body: _hints-script.html
    toc: true
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    page-layout: article
    anchor-sections: true
    code-line-numbers: true
    highlight-style: github
---


::: {.path-crumb}
M1 Economics / Intro to Programming / <span class="here">recap-rstudio-project</span>
:::

::: {.hero-tree}
project.Rproj/<br>
&nbsp;&nbsp;<span class="dir">data/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">raw/</span><span class="cursor"></span>
:::

::: {.objective}
#### Objective of this recap

Before starting any analysis, you need a properly organized RStudio project. This short recap walks you through:

1. creating a new RStudio project,
2. building the standard folder structure used throughout the course.
:::

## Setup

Adopt the tree structure presented below for every new project you create in this course. Note that launching your session via `project.Rproj` sets your active working directory to the folder containing it.

::: {.hero-tree style="display:block;"}
<span class="dir">Root/</span><br>
&nbsp;&nbsp;<span class="dir">data/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">raw/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">tmp/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">out/</span><br>
&nbsp;&nbsp;<span class="dir">functions/</span><br>
&nbsp;&nbsp;<span class="dir">scripts/</span><br>
&nbsp;&nbsp;<span class="dir">notebooks/</span><br>
&nbsp;&nbsp;<span class="dir">figs/</span><br>
&nbsp;&nbsp;<span class="dir">tables/</span><br>
&nbsp;&nbsp;<span class="dir">references/</span><br>
&nbsp;&nbsp;<span class="file">project.Rproj</span>
:::

::: {.qcard}
**Step 1. Open RStudio**

Launch RStudio on your machine.
:::

::: {.qcard}
**Step 2. Create a new project**

* `File` → `New Project...`
* Choose **New Directory**
* Name your project (e.g. `session-2`)
:::

::: {.qcard}
**Step 3. Create the project structure**

In the R console, run:

```{r}
#| eval: false
#| echo: true
dir.create("data")
dir.create("data/raw")
dir.create("data/tmp")
dir.create("data/out")
dir.create("functions")
dir.create("scripts")
dir.create("notebooks")
dir.create("figs")
dir.create("tables")
dir.create("references")
```
:::

::: {.tip}
#### Check your setup

Once the folders are created, you can confirm everything is in place directly from the R console:

```{r}
#| eval: false
getwd()      # should point to the folder containing project.Rproj
list.files() # should list all the folders you just created
```
:::

<div class="text-center">
  <a href="index.html" class="btn btn-success">
    {{< fa house >}} Course homepage
  </a>
</div>

---

*Introduction to Programming for Data Analysis — Master 1 in Economics*
