---
title: "Tutorial #3. Linear regressions and statistical tests"
subtitle: "Introduction to programming for data analysis"
author:
  - name: Ulrich Aiounou
    email: dimitri.aiounou@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Ewen Gallic
    email: ewen.gallic@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Kla Kouadio
    email: kla.kouadio@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Pierre Michel
    email: pierre.michel@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
format:
  html:
    css: styles.css
    include-in-header: _fonts.html
    include-after-body: _hints-script.html
    include-before-body: _topbar.html
    toc: true
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    page-layout: article
    anchor-sections: true
    code-line-numbers: true
    highlight-style: github
navbar:
  right:
    - text: "Home"
      href: "../"
---



::: {.path-crumb}
M1 Economics / Intro to Programming / <span class="here">tutorial-3</span>
:::

::: {.warning}

## Disclaimer

This HTML page was generated with Claude Sonnet 5 based on exercises we developed from a LaTeX document.

If you would prefer to work with a more traditional PDF version, you can download it here: [Download the PDF](tutorial-3.pdf).

Part of the content was created by Pierre Michel (AMSE) and Morgan Raux (AMSE) who kindly shared their work.

:::

This tutorial focuses on regression models and statistical hypothesis testing. After completing Tutorial #2, you should be able to create basic data visualizations and compute summary statistics. Building on these skills, you are now ready to run regression analyses and apply statistical modeling techniques.

In Tutorial #1, you downloaded datasets from Eurostat, merged them into a single dataset in R, and exported the result as a CSV file.

::: {.objective}
#### Objectives of the tutorial

In this tutorial, you will use the dataset from exercise 2 of Tutorial #1 to:

1. create visualizations to examine the relationship between the two main variables (per capita GDP and overall life satisfaction),
2. estimate regression models, including both linear and nonlinear specifications for panel data.
:::

```{r}
#| include: false
library(readr)
library(dplyr)
library(ggplot2)
```

::: {.tip}

#### Did not finish tutorial 1?

Didn't get to finish combining the data in Tutorial 1? Grab the ready-to-use file below and save it to the `data/out/` folder of your project.

- <a href="data/out/gdp_lifesat.csv" class="btn btn-light btn-sm" data-toggle="dropdown" aria-haspopup="true" aria-expanded="false" download>{{< fa download >}} gdp_lifesat.csv</a>

:::

::: {.qcard}
**Q1.** Import the dataset that you saved at the end of exercise 2 of Tutorial #1.

```{r}
#| eval: false
library(tidyverse)
library(stargazer)
tb <- ...
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
The file lives in your `data/out/` folder — build the relative path from your project root.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
library(readr)
# Import data path from your project repository structure
tb <- read_csv("data/out/gdp-lifesat.csv")
tb
```
:::
:::

## Scatterplots

A scatterplot visually represents the relationship between two variables. In the current dataset, it is useful to examine how per capita GDP relates to overall life satisfaction.

### GDP and life satisfaction

::: {.qcard}
**Q2.** Create a scatterplot to display the relationship between per capita GDP and overall life satisfaction. Label the x-axis "Per capita GDP" and the y-axis "Overall Life Satisfaction". Use the title "Per capita GDP vs. Overall Life Satisfaction" for the graph.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Map `x = gdp` and `y = life_satisf` inside `aes()`, then use `geom_point()`.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
ggplot(data = tb, mapping = aes(x = gdp, y = life_satisf)) +
    geom_point(alpha = 0.6) +
    labs(
        x = "Per capita GDP",
        y = "Overall Life Satisfaction",
        title = "Per capita GDP vs. Overall Life Satisfaction"
    ) +
    theme_minimal()
```
:::
:::

::: {.qcard}
**Q3.** Add a linear regression line to the scatterplot.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
`geom_smooth()` adds a fitted trend line; set `method = "lm"` for a straight linear fit, and `se = TRUE` to show the confidence band.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
ggplot(data = tb, mapping = aes(x = gdp, y = life_satisf)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", color = "blue", se = TRUE) +
    labs(
        x = "Per capita GDP",
        y = "Overall Life Satisfaction",
        title = "Per capita GDP vs. Overall Life Satisfaction (Linear)"
    ) +
    theme_minimal()
```
:::
:::

::: {.qcard}
**Q4.** Add a quadratic curve to the scatterplot.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Still `method = "lm"`, but change the `formula` argument to include a squared term: `y ~ x + I(x^2)`.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
ggplot(data = tb, mapping = aes(x = gdp, y = life_satisf)) +
    geom_point(alpha = 0.6) +
    geom_smooth(
        method = "lm", formula = y ~ x + I(x^2), color = "red", se = TRUE
    ) +
    labs(
        x = "Per capita GDP",
        y = "Overall Life Satisfaction",
        title = "Per capita GDP vs. Overall Life Satisfaction (Quadratic)"
    ) +
    theme_minimal()
```
:::
:::

### Changes in GDP and life satisfaction

::: {.qcard}
**Q5.** Create two new columns, `gdp_yoy` and `life_satisf_yoy`, which represent the within-country year-on-year changes in GDP and life satisfaction, respectively.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn hint" data-step="2" disabled><span class="dot"></span>Hint 2</button>
<button class="step-btn sol" data-step="3" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
This is panel data: to compute a year-on-year change correctly, you must first sort observations chronologically **within each country**.
:::

::: {.reveal .hint-box data-reveal="2"}
`arrange(country, year)` followed by `group_by(country)`, then use `lag()` inside `mutate()` to access the previous year's value.
:::

::: {.reveal .sol-box data-reveal="3"}
Because this is panel data, we must sort chronologically and group by country before calculating changes to avoid mixed metrics between borders:

```{r}
tb <- tb |>
    arrange(country, year) |>
    group_by(country) |>
    mutate(
        gdp_yoy = gdp - lag(gdp),
        life_satisf_yoy = life_satisf - lag(life_satisf)
    ) |>
    ungroup()
tb
```
:::
:::

::: {.qcard}
**Q6.** Create a scatterplot to illustrate the relationship between `gdp_yoy` and `life_satisf_yoy`.

- Place the year-on-year variation in life satisfaction on the y-axis and label it "Yearly Change in Life Satisfaction".
- Place the year-on-year variation in per capita GDP on the x-axis and label it "Yearly Change in Per Capita GDP".
- Title the graph "Relationship Between Changes in GDP and Life Satisfaction".
- Add a linear regression line to the scatterplot.

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
ggplot(data = tb, mapping = aes(x = gdp_yoy, y = life_satisf_yoy)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "darkgreen") +
    labs(
        x = "Yearly Change in Per Capita GDP",
        y = "Yearly Change in Life Satisfaction",
        title = "Relationship Between Changes in GDP and Life Satisfaction"
    ) +
    theme_minimal()
```
:::
:::

## OLS Regression

::: {.qcard}
**Q7.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
#| output: false
model_lm <- lm(life_satisf ~ gdp, data = tb)
summary(model_lm)
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
The `lm()` command calculates a standard pooled ordinary least squares (OLS) linear model mapping the effect of independent variable `gdp` onto dependent metric `life_satisf`. Calling `summary()` prints the descriptive evaluation properties: regression estimates ($\hat{\beta}$ parameters), standard errors, $t$-statistics, associated $p$-values, and overall model diagnostics ($R^2$ variance performance).
```{r}
#| echo: false
summary(model_lm)
```

:::
:::

::: {.qcard}
**Q8.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
#| eval: false
library(stargazer)
stargazer(model_lm, type = "latex", out = "tables/model_lm.tex")
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
The function converts the raw statistical model object metrics directly into a formatted LaTeX tabular canvas, printing and writing the compiled outputs to the external filepath location: `tables/model_lm.tex`.

```{r}
#| echo: false
#| message: false
#| warning: false
library(stargazer)
stargazer(model_lm, type = "latex", out = "tables/model_lm.tex")
```

:::
:::

### Fixed effects regression

::: {.qcard}
**Q9.** Extend the previous regression model by including country fixed effects to account for unobserved, time-invariant differences across countries.

```{r}
#| eval: false
model_country_fe <- ...
summary(model_country_fe)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Add `country` as a right-hand-side term, wrapped in `factor()` so it is treated as a categorical variable (a set of dummies) rather than a numeric one.
:::

::: {.reveal .sol-box data-reveal="2"}
We append country variables as dummy indices using the `factor()` wrapper:

```{r}
model_country_fe <- lm(life_satisf ~ gdp + factor(country), data = tb)
summary(model_country_fe)
```
:::
:::

::: {.qcard}
**Q10.** Extend the country fixed-effects model by adding year fixed effects to control for common shocks across all countries in a given year.

```{r}
#| eval: false
model_country_time_fe <- ...
summary(model_country_time_fe)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Same idea as Q9, but add a second `factor()` term for `year` alongside `factor(country)`.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
model_country_time_fe <- lm(
    life_satisf ~ gdp + factor(country) + factor(year), 
    data = tb
)
summary(model_country_time_fe)
```
:::
:::

::: {.qcard}
**Q11.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
#| eval: false
stargazer(
    model_lm, model_country_fe, model_country_time_fe,
    type = "latex",
    out = "tables/regression_results.tex",
    title = "Correlation between per capita GDP and life satisfaction",
    column.labels = c("Baseline OLS Reg.", "Country FE", "Country and Year FE"),
    covariate.labels = c("per capita GDP"),
    add.lines = list(
        c('Country FE', '-', 'Yes', 'Yes'),
        c('Year FE', '-', '-', 'Yes')
    )
)
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
This block merges the baseline, country FE, and two-way FE models into a single side-by-side LaTeX compilation table. It replaces the long list of individual dummy coefficients with neat summary rows using custom indicators in `add.lines`, saves the resulting output inside `tables/regression_results.tex`, and updates the column titles appropriately.

```{r}
#| echo: false
#| message: false
#| warning: false
stargazer(
    model_lm, model_country_fe, model_country_time_fe,
    type = "latex",
    out = "tables/regression_results.tex",
    title = "Correlation between per capita GDP and life satisfaction",
    column.labels = c("Baseline OLS Reg.", "Country FE", "Country and Year FE"),
    covariate.labels = c("per capita GDP"),
    add.lines = list(
        c('Country FE', '-', 'Yes', 'Yes'),
        c('Year FE', '-', '-', 'Yes')
    )
)
```


:::
:::

## Non-linear regression

We herein consider non-linear regression models with a dummy as dependent variable, such as probit and logit regression models.

::: {.qcard}
**Q12.** Compute the median of life satisfaction.

```{r}
#| eval: false
median_life_satisf <- ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
median_life_satisf <- median(tb$life_satisf, na.rm = TRUE)
median_life_satisf
```
:::
:::

::: {.qcard}
**Q13.** In the dataset, create a binary variable named `life_satisf_above_med`, which takes the value 1 if life satisfaction is above the median and 0 otherwise.

```{r}
#| eval: false
tb <- ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
tb <- tb |>
    mutate(
        life_satisf_above_med = ifelse(
            life_satisf > median_life_satisf, 
            1,
            0
        )
    )
tb
```
:::
:::

::: {.qcard}
**Q14.** Run a **probit** regression of `life_satisf_above_med` on `gdp`, and display the summary of the results.

```{r}
#| eval: false
model_probit <- ...
summary(model_probit)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
With a binary dependent variable, `lm()` is no longer appropriate — use `glm()` with `family = binomial(link = "probit")`.
:::

::: {.reveal .sol-box data-reveal="2"}
We utilize generalized linear model modeling via `glm()` specifying link definitions:

```{r}
model_probit <- glm(
    life_satisf_above_med ~ gdp, 
    data = tb,
    family = binomial(link = "probit")
)
summary(model_probit)
```
:::
:::

::: {.qcard}
**Q15.** Run a **logit** regression of `life_satisf_above_med` on `gdp`, and display the summary of the results.

```{r}
#| eval: false
model_logit <- ...
summary(model_logit)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Same as Q14, but change the link function to `"logit"`.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
model_logit <- glm(
    life_satisf_above_med ~ gdp, 
    data = tb,
    family = binomial(link = "logit")
)
summary(model_logit)
```
:::
:::

## Statistical tests

::: {.qcard}
**Q16.** Compute the median of per capita GDP.

```{r}
#| eval: false
median_gdp <- ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
median_gdp <- median(tb$gdp, na.rm = TRUE)
median_gdp
```
:::
:::

::: {.qcard}
**Q17.** In the dataset, create a binary variable named `gdp_above_med`, which takes the value 1 if per capita GDP is above the median and 0 otherwise.

```{r}
#| eval: false
tb <- ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
tb <- tb |>
    mutate(gdp_above_med = ifelse(gdp > median_gdp, 1, 0))
tb
```
:::
:::

::: {.qcard}
**Q18.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
group1 <- tb |>
    filter(gdp_above_med == 0) |>
    pull(life_satisf)
    
group2 <- tb |>
    filter(gdp_above_med == 1) |>
    pull(life_satisf)
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
These two instructions split the dataset into two independent numeric vectors: `group1` contains the life satisfaction values for countries with below-median GDP, and `group2` contains the life satisfaction values for countries with above-median GDP. `pull()` extracts a single column as a plain vector, rather than as a one-column data frame.
:::
:::

::: {.qcard}
**Q19.** Perform a Student's t-test to compare the mean life satisfaction between observations with GDP above the median and those with GDP at or below the median.

```{r}
#| eval: false
t_test_result <- ...
print(t_test_result)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Use `t.test()` on the two vectors created in Q18, with `alternative = "two.sided"` since we have no prior expectation on the direction of the difference.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
t_test_result <- t.test(group2, group1, alternative = "two.sided")
print(t_test_result)
```
:::
:::

::: {.qcard}
**Q20.** Extract the test statistic and the p-value from the results of the t-test.

```{r}
#| eval: false
test_statistic <- ...
p_value <- ...
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
The object returned by `t.test()` is a list — access its named elements with the `$` operator, just like a column of a data frame.
:::

::: {.reveal .sol-box data-reveal="2"}
Using list element indexing on the returned object properties:

```{r}
test_statistic <- t_test_result$statistic
p_value <- t_test_result$p.value
test_statistic
p_value
```
:::
:::

::: {.qcard}
**Q21.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
latex_table <- paste0(
    "\\begin{tabular}{l c}\n",
    "\\hline\n",
    "& Difference in life satisfaction across groups" , 
    " \\\\ \n",
    "Test Statistic & ", 
    format(test_statistic, digits = 3), 
    " \\\\ \n",
    "P-value & ", 
    format.pval(p_value, digits = 3), 
    " \\\\ \n",
    "\\hline\n",
    "\\end{tabular}"
)
file_path <- "tables/table_results_t_test_1_v1.tex"
writeLines(latex_table, file_path)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
`paste0()` glues character strings together with no separator — here, it is building a LaTeX `tabular` environment line by line, as a single block of text.
:::

::: {.reveal .sol-box data-reveal="2"}
This block manually builds a small LaTeX table (a `tabular` environment with two rows: the test statistic and the p-value) as a character string, using `paste0()` to concatenate the LaTeX syntax with the formatted numeric results. `format()` and `format.pval()` control the number of significant digits displayed. `writeLines()` then writes this character string to the file `tables/table_results_t_test_1_v1.tex`, one line at a time. This is a manual alternative to `stargazer()`, useful for results (like a t-test) that are not regression model objects.

```{r}
#| echo: false
cat(latex_table)
```
:::
:::


<div class="text-center">
  <a href="tutorial-2.html" class="btn btn-success">
    {{< fa arrow-left >}} Tutorial 2
  </a>
</div>

---

*Introduction to Programming for Data Analysis — Master 1 in Economics*
