---
title: "Tutorial #2. Data visualization and summary statistics"
subtitle: "Introduction to programming for data analysis"
author:
  - name: Ulrich Aiounou
    email: dimitri.aiounou@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Ewen Gallic
    email: ewen.gallic@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Kla Kouadio
    email: kla.kouadio@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
  - name: Pierre Michel
    email: pierre.michel@univ-amu.fr
    affiliations:
      - name: AMSE, Aix Marseille University
format:
  html:
    css: styles.css
    include-in-header: _fonts.html
    include-before-body: _topbar.html
    include-after-body: _hints-script.html
    toc: true
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    page-layout: article
    anchor-sections: true
    code-line-numbers: true
    highlight-style: github
---


::: {.path-crumb}
M1 Economics / Intro to Programming / <span class="here">tutorial-2</span>
:::

::: {.hero-tree}
tutorial-2/<br>
&nbsp;&nbsp;<span class="dir">data/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">out/</span> <span class="file">gdp-lifesat.csv</span><br>
&nbsp;&nbsp;<span class="dir">scripts/</span><span class="cursor"></span>
:::

::: {.warning}

## Disclaimer

This HTML page was generated with Claude Sonnet 5 based on exercises we developed from a LaTeX document.

If you would prefer to work with a more traditional PDF version, you can download it here: [Download the PDF](tutorial-2.pdf).

Part of the content was created by Pierre Michel (AMSE) and Morgan Raux (AMSE) who kindly shared their work.

:::


In the previous tutorial, you downloaded datasets from Eurostat, merged them into a single dataset in R, and exported the result as a CSV file.

::: {.objective}
#### Objectives of the tutorial

In this tutorial, you will work with the dataset from the second exercise of Tutorial #1 to:

1. explore it through basic summary statistics,
2. create your first data visualizations in R using `ggplot2`.
:::

::: {.tip}

#### Did not finish tutorial 1?

Didn't get to finish combining the data in Tutorial 1? Grab the ready-to-use file below and save it to the `data/out/` folder of your project.

- <a href="data/out/gdp_lifesat.csv" class="btn btn-light btn-sm" data-toggle="dropdown" aria-haspopup="true" aria-expanded="false" download>{{< fa download >}} gdp_lifesat.csv</a>

:::


::: {.tip}
#### 💡 Clean session

If you need to have a clean session in RStudio, first remove all the objects in memory by evaluating the following instruction:

```{r}
#| eval: false
rm(list = ls())
```

Then, on the menu "Session", click on "Restart R".
:::

## Setup

Prepare a project for the second tutorial. Adopt the tree structure presented below. Note that launching your session via `project.Rproj` sets your active working directory to `tutorial-2/`.

::: {.hero-tree style="display:block;"}
tutorial-2/<br>
&nbsp;&nbsp;<span class="dir">data/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">raw/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">tmp/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;<span class="dir">out/</span><br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span class="file">gdp-lifesat.csv</span><br>
&nbsp;&nbsp;<span class="dir">scripts/</span><br>
&nbsp;&nbsp;<span class="dir">figs/</span><br>
&nbsp;&nbsp;<span class="dir">tables/</span><br>
&nbsp;&nbsp;<span class="dir">references/</span><br>
&nbsp;&nbsp;<span class="file">README.md</span><br>
&nbsp;&nbsp;<span class="file">project.Rproj</span>
:::

```{r}
#| include: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

::: {.qcard}
**Q1.** Import the dataset you exported at the end of Tutorial #1 (`gdp-lifesat.csv`, assumed to be copied into your data entry structure).

```{r}
#| eval: false
tb <- read_csv(...)
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
The file lives in your `data/out/` folder — build the relative path from your project root (`tutorial-2/`).
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
library(readr)
# Assuming the file was moved to the data output folder of the project root
tb <- read_csv("data/out/gdp-lifesat.csv")
tb
```
:::
:::

## Plot Distributions

::: {.qcard}
**Q2.** Read the following code, evaluate it, and explain what it does.

```{r}
#| output: false
p_density <- ggplot(
  data = tb, 
  mapping = aes(x = gdp)
) +
  geom_density(fill = "blue", alpha = 0.1) +
  labs(
    x = "GDP", 
    y = "Density",
    title = "Density Plot of GDP"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )
p_density
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
This block initialises a graphic via `ggplot2` using the `tb` dataset. It maps the continuous metric `gdp` to the x-axis and draws an empirical density distribution via `geom_density()`, filled with light blue (10% opacity). It adds clean labels and applies a minimal theme. The background grid lines are removed.
```{r}
#| echo: false
p_density
```

:::
:::

::: {.qcard}
**Q3.** Using the `ggsave()` function, export this graph as a PDF file in the `figs` folder of your project environment.

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
ggsave("figs/density_plot_gdp.pdf", plot = p_density, width = 7, height = 5)
```
:::
:::

::: {.qcard}
**Q4.** Plot the distribution of the variable measuring life satisfaction using a histogram:

- Label the x-axis as "Overall life satisfaction".
- Remove the y-axis title.
- Add the plot title "Histogram of Overall life satisfaction".

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Use `geom_histogram()`, and set `y = NULL` inside `labs()` to remove an axis title.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
ggplot(data = tb, mapping = aes(x = life_satisf)) +
  geom_histogram(bins = 15, fill = "darkgray", color = "white") +
  labs(
    x = "Overall life satisfaction",
    y = NULL, # Removes the y-axis title
    title = "Histogram of Overall life satisfaction"
  ) +
  theme_minimal()
```
:::
:::

## Summary Statistics

::: {.qcard}
**Q5.** Examine and evaluate the following instructions. Then, explain what they do.

```{r}
#| output: false
summary_stats_base <- tb |>
  select(gdp, life_satisf) |>
  summary()

print(summary_stats_base)
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
This pipeline of instructions extracts the features `gdp` and `life_satisf` from the data frame `tb` and passes them directly into R's base summary function `summary()`. This returns the descriptive quantities (Minimum, 1st Quartile, Median, Mean, 3rd Quartile, Maximum) alongside missing item flags (`NA`s) per variable.
```{r}
#| echo: false
print(summary_stats_base)
```

:::
:::

::: {.qcard}
**Q6.** Create a table of summary statistics for the variables `gdp` and `life_satisf`. The table should include the following statistics: mean, standard deviation, minimum, maximum. Round all values to two decimal places. The resulting table must have two rows (one for `gdp`, one for `life_satisf`). Name this table `summary_stats`.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn hint" data-step="2" disabled><span class="dot"></span>Hint 2</button>
<button class="step-btn sol" data-step="3" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
To get one row per variable, first reshape the two columns into a long format with `pivot_longer()`.
:::

::: {.reveal .hint-box data-reveal="2"}
Then `group_by(var)` and `summarise()` with `mean()`, `sd()`, `min()`, `max()`, each wrapped in `round(..., 2)`.
:::

::: {.reveal .sol-box data-reveal="3"}
Using tidyverse strategies to explicitly construct long summaries:

```{r}
library(tidyr)

summary_stats <- tb |>
  select(gdp, life_satisf) |>
  pivot_longer(
    cols = everything(), names_to = "var", values_to = "val"
  ) |>
  group_by(var) |>
  summarise(
    mean = round(mean(val, na.rm = TRUE), 2),
    sd   = round(sd(val, na.rm = TRUE), 2),
    min  = round(min(val, na.rm = TRUE), 2),
    max  = round(max(val, na.rm = TRUE), 2)
  )
summary_stats
```
:::
:::

::: {.qcard}
**Q7.** Examine and evaluate the following instruction. Then, explain what it does.

```{r}
#| output: false
names(summary_stats) <- c("var", "mean", "sd", "min", "max")
summary_stats
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
It replaces the names of the columns in the `summary_stats` summary table with those provided in the vector (`c()`). Here the names happen to already match, so the table is unchanged — but this instruction is what you would use to rename columns coming out of a pipeline with less convenient default names.
```{r}
#| echo: false
summary_stats
```

:::
:::

::: {.qcard}
**Q8.** Print the content of the object `summary_stats` as a LaTeX-formatted table. To do so, use the `{stargazer}` package. Adapt the following code:

```{r}
#| eval: false
install.packages("stargazer")
library(stargazer)
?stargazer
summary_stats_latex <- stargazer(...)
```

Save the LaTeX table into a file named `summary_stats.tex` inside your output data subfolder.

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::


::: {.reveal .sol-box data-reveal="2"}
```{r}
#| eval: true
library(stargazer)

stargazer(
  summary_stats,
  summary = FALSE, 
  type = "latex", 
  out = "data/out/summary_stats.tex"
)
```


:::
:::

## Rank countries by GDP in 2023

::: {.qcard}
**Q9.** Create an object called `tb_2023` that contains only the rows of the dataset where the year is 2023.

```{r}
#| eval: false
tb_2023 <- tb |>
    ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
tb_2023 <- tb |> 
  filter(year == 2023)
tb_2023
```
:::
:::

::: {.qcard}
**Q10.** Examine and evaluate the following instruction. Then, explain what it does.

```{r}
#| output: false
p <- ggplot(
  data = tb_2023, 
  mapping = aes(x = reorder(country, gdp), y = gdp)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "GDP by Country in 2023",
    x = "Country",
    y = "GDP"
  ) +
  theme_minimal()
p
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
This instruction prints a sorted horizontal bar chart. `reorder(country, gdp)` orders the categorical country axis sequentially by their corresponding value metric (that of variable `gdp`). `geom_bar(stat = "identity")` maps raw values directly to lengths, and `coord_flip()` rotates the layout by 90 degrees, swapping the x- and y-axes.
```{r}
#| echo: false
p
```
:::
:::

::: {.qcard}
**Q11.** Create an object called `data_filtered` that contains only the rows of the dataset where the year is either 2014 or 2023.

```{r}
#| eval: false
data_filtered <- ...
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Use `%in%` with a two-element vector, rather than two chained `==` conditions.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
data_filtered <- tb |> 
  filter(year %in% c(2014, 2023))
data_filtered
```
:::
:::

::: {.qcard}
**Q12.** Examine and evaluate the following instruction. Then, explain what it does.

```{r}
#| eval: false
data_filtered |>
  arrange(year, desc(gdp))
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
It sorts rows first chronologically by `year`, then sorts the observations in descending order of cross-sectional `gdp` within each temporal block.
```{r}
#| echo: false
data_filtered |>
  arrange(year, desc(gdp))
```
:::
:::

::: {.qcard}
**Q13.** Create a barplot ranking countries in decreasing order of GDP, with two panels: the left panel for year 2014, the right panel for year 2023. The plot should have:

1. the x-axis titled "GDP",
2. no y-axis title,
3. the main title "GDP by Country in 2014 and 2023".

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn sol" data-step="2" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Use `facet_wrap(~year)` on top of the same bar chart recipe as Q10, then swap which axis gets the `NULL` label.
:::

::: {.reveal .sol-box data-reveal="2"}
```{r}
ggplot(data = data_filtered, mapping = aes(x = reorder(country, gdp), y = gdp)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~year) + # Splitting into panels by year
  labs(
    x = NULL,
    y = "GDP",
    title = "GDP by Country in 2014 and 2023"
  ) +
  theme_minimal()
```
:::
:::

## Rank countries by change in GDP between 2014 and 2023

::: {.qcard}
**Q14.** Create a new variable called `change_gdp` that measures the within-country change in GDP between 2014 and 2023.

```{r}
#| eval: false
tb <- tb |> ...
```

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn hint" data-step="2" disabled><span class="dot"></span>Hint 2</button>
<button class="step-btn sol" data-step="3" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
This is a within-country comparison across years, so you need one row per country: reshape the data with `pivot_wider()` so that each year's GDP becomes its own column.
:::

::: {.reveal .hint-box data-reveal="2"}
`pivot_wider(names_from = year, values_from = gdp, names_prefix = "gdp_")` gives you columns `gdp_2014` and `gdp_2023`. Then simply subtract them with `mutate()`.
:::

::: {.reveal .sol-box data-reveal="3"}
To extract delta values properly across years from wide or long frames:

```{r}
tb_change <- tb |>
  filter(year %in% c(2014, 2023)) |>
  select(country, year, gdp) |>
  pivot_wider(names_from = year, values_from = gdp, names_prefix = "gdp_") |>
  mutate(change_gdp = gdp_2023 - gdp_2014)
tb_change
```
:::
:::

::: {.qcard}
**Q15.** Examine and evaluate the following instruction. Then, explain what it does.

```{r}
#| output: false
tb_2023 <- tb |> filter(year == 2023)
tb_2023
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
It isolates cross-sectional observations from 2023 from the data frame `tb` and assigns the resulting data frame in an object named `tb_2023`.
```{r}
#| echo: false
tb_2023
```
:::
:::

::: {.qcard}
**Q16.** Create a barplot ranking countries in decreasing order of their change in GDP (between 2014 and 2023). The plot should have:

- the x-axis titled "Change in GDP",
- no y-axis title,
- the main title "Change in GDP by Country between 2014 and 2023".

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
ggplot(data = tb_change, mapping = aes(x = reorder(country, change_gdp), y = change_gdp)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Change in GDP",
    title = "Change in GDP by Country between 2014 and 2023"
  ) +
  theme_minimal()
```
:::
:::

## Evolution of GDP in France and Italy

::: {.qcard}
**Q17.** Create an object called `tb_france_italy` that contains rows of the dataset `tb` where the country is either "France" or "Italy".

```{r}
#| eval: false
tb_france_italy <- tb |> ...
```

::: {.cascade}
<button class="step-btn sol" data-step="1"><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .sol-box data-reveal="1"}
```{r}
tb_france_italy <- tb |> 
  filter(country %in% c("FR", "IT"))
tb_france_italy
```
:::
:::

::: {.qcard}
**Q18.** Using the dataset `tb_france_italy`, create a line plot showing the year-by-year evolution of GDP for France and Italy (one line per country). Include a legend to indicate which line corresponds to each country.

Bonus: use blue (`#0055A4`) for France and green (`#009246`) for Italy. Place the legend below the graph. Title the y-axis "GDP".

::: {.cascade}
<button class="step-btn hint" data-step="1"><span class="dot"></span>Hint 1</button>
<button class="step-btn hint" data-step="2" disabled><span class="dot"></span>Hint 2</button>
<button class="step-btn sol" data-step="3" disabled><span class="dot"></span>Show solution</button>
<div class="progress-track"><div class="progress-fill"></div></div>
:::

::: {.reveal .hint-box data-reveal="1"}
Map `color = country` (and `group = country`) inside `aes()` so `ggplot2` draws one line per country and builds the legend automatically.
:::

::: {.reveal .hint-box data-reveal="2"}
For custom colors, use `scale_color_manual(values = c("FR" = ..., "IT" = ...))`. For the legend position, set `theme(legend.position = "bottom")`.
:::

::: {.reveal .sol-box data-reveal="3"}
```{r}
ggplot(data = tb_france_italy, mapping = aes(x = year, y = gdp, color = country, group = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_manual(values = c("FR" = "#0055A4", "IT" = "#009246")) +
  labs(
    x = "Year",
    y = "GDP",
    title = "Evolution of GDP in France and Italy"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
:::
:::

<div class="text-center">
  <a href="tutorial-1.html" class="btn btn-success">
    {{< fa arrow-left >}} Tutorial 1
  </a>
  <a href="tutorial-3.html" class="btn btn-success">
    {{< fa arrow-right >}} Tutorial 3
  </a>
</div>

---

*Introduction to Programming for Data Analysis — Master 1 in Economics*
